### 2014-10-29 motivation.tex
Spacing edits to motivation, contract-language. More text in transactions.
Added transaction section
Added RDT example in motivation

\section{System Model}
\label{sec:sysmod}

\begin{figure}
\centering
\includegraphics[width=0.75\columnwidth]{Figures/SystemModel}
\caption{Quelea system model.}
\label{fig:sysmod}
\end{figure}

Figure~\ref{fig:sysmod} provides a schematic diagram of our system model. The
distributed store is composed of a collection of \emph{replicas}, each of which
stores a set of \emph{objects} ($x,y,\ldots$). We assume that every object is
replicated at every replica in the store. The state of an object at any replica
is the set of all updates (\emph{effects}) performed on the object. For
example, the state of $x$ at replica 1 is the set composed of effects $w^x_1$
and $w^x_2$.

Each object is associated with a set of \emph{operations}. The clients interact
with the store by invoking operations on objects. The sequence of operations
invoked by a particular client on the store is called a \emph{session}. The
data store is typically accessed by a large number of clients (and hence
sessions) concurrently. Importantly, the clients are oblivious to which replica
an operation is applied to; the data store may choose to route the operation to
the closest replica for minimizing latency, or route it to a different replica
in order to balance the load in the system. For example, the operations
\emph{foo} and \emph{bar} invoked by the same session on the same object, might
end up being applied to different replicas because replica 1 (to which
\emph{foo} was applied) might be unreachable when the client invokes
\emph{bar}.

When \emph{foo} is invoked on a object $x$ with arguments \emph{arg}$_1$ at
replica 1, it simply \emph{reduces} over the current set of effects at that
replica on that object ($w^x_1$ and $w^x_2$), produces a result $v1$ that is
sent back to the client, and emits a \emph{single} new effect $w^x_4$ that is
appended to the state of $x$ at replica 1. Thus, every operation is evaluated
over a \emph{snapshot} of the state of the object on which it is invoked. In
this case, we say that the effect $w^x_4$ \emph{witnesses} $w^x_1$ and $w^x_2$.
Alternatively, $w^x_1$ and $w^x_2$ are \emph{visible} to $w^x_4$, written
logically as $\vis{w^x_1}{w^x_4} \wedge \vis{w^x_2}{w^x_4}$, where $\visZ$ is
the visibility relation between effects. Visibility is an irreflexive (an
operation cannot see its own effect) and asymmetric relation, and only relates
effects produced by operations on the same object. Executing a read-only
operation is similar except that no new effects are produced.

The effect added to a particular replica is asynchronously sent to other
replicas. We assume that \emph{eventually} an effect applied at one replica
will be merged into all other replicas. Two effects $w^x_4$ and $w^x_5$ that
arise from the same session are said to be in \emph{session order} (written
logically as $\so{w^x_4}{w^x_5}$). Session order is an irreflexive, transitive
relation. The effects $w^x_4$ and $w^x_5$ arising from operations applied to
the same object $x$ are said to be under the \emph{same object} relation,
written $\sameobj{w^x_4}{w^x_5}$. Finally, we can associate every effect with
the operation that generated the effect with the help of a relation $\operZ$.
In the current example, $\oper{w^x_4}{foo}$ and $\oper{w^x_5}{bar}$ hold. For
simplicity, we assume all operation names across all object types are distinct.

This model admits all the inconsistencies associated with eventual consistency.
The goal of this work is to identify the precise consistency level for each
operation such that application-level constraints are not violated. In the next
section, we will concretely describe the challenges associated with
constructing a consistent bank account on top of an eventually consistent data
store. Subsequently, we will illustrate how our contract and specification
language, armed with the primitive relations $\visZ$, $\soZ$, $\sameobjZ$ and
$\operZ$, mitigates these challenges.

@@

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. Let us assume that the bank account object
provides three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance},
with the assumption that the withdraw fails if the account has insufficient
balance. Our goal is to choose the correct consistency level for each of these
operations such that (1) the balance never drops below zero and (2) the
\cf{getBalance} operation never incorrectly returns a negative balance. To be
able to do this, we need to consider the anomalies that could arise under
eventual consistency.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification over \cf{withdraw}:
\vspace{-.5em}
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}
\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

\noindent For \cf{getBalance}, we construct the following contract:
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}
\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

@@

Since there are no restrictions on when and how a \cf{deposit} operation can
execute, its contract is simply $\true$.

>>>>

@@

Consider how we might implement a highly available bank account on top of an
eventually consistent data store, with the \emph{integrity} constraint that the
balance should never drop below 0. We will begin by implementing the bank
account replicated data type (RDT) in \name, and then describe the mechanisms
to obtain the desired correctness guarantees.

\subsection{RDT specification}

A key novelty	in \name is that it allows the addition of new RDTs to the store,
immensely improving the programmability of data stores. The programmer no
longer has to coerce the application logic to utilize the store provided data
types. In addition, \name treats the convergence semantics of the data type
separately from the consistency properties. The separation of concern allows
the programmer to \emph{operationally} reason about conflict resolution, while
permitting \emph{declarative} reasoning for consistency. As mentioned in the
previous section, RDTs are described in terms of operations on a RDT object,
where each operation is described as a reduction over the state of the object
i.e., the list of effects. Let us assume that the bank account object provides
three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the
assumption that the withdraw fails if the account has insufficient balance.
Every operation in \name is of the following type, written in Haskell syntax:

\begin{codehaskell}
type Operation = (Effect e, Arg a, Res r)
               => [e] -> a -> (r, Maybe e)
\end{codehaskell}

\noindent It takes a list of effects (the \emph{context} for the operation), an
argument and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations is given
below:

\begin{codehaskell}
data Acc = Deposit_ Int | Withdraw_ Int | GetBalance_

getBalance :: [Acc] -> () -> (Int, Maybe Acc)
getBalance ctxt _ =
  let res = sum [x | Deposit_ x <- ctxt]
						- sum [x | Withdraw_ x <- ctxt]
	in (res, Nothing)

deposit :: [Acc] -> Int -> ((), Maybe Acc)
deposit _ amt = (amt, Just $ Deposit_ amt)

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 $ getBalance ctxt () >= v
  then (True, Just $ Withdraw_ v)
	else (False, Nothing)
\end{codehaskell}

The datatype \cf{Acc} represents the effect type for the bank account. The
context of the operations is a snapshot of the state of the object at some
replica. In this sense, every operation on the RDT is atomic, and thus
permitting sequential reasoning for implementing eventually consistent data
types. We have implemented a large corpus of RDTs for realistic benchmarks
including shopping carts, auction and micro-blogging sites in few tens of lines
of code.

\subsection{Anomalies under Eventual Consistency}

Our goal is to choose the correct consistency level for each of the bank
account operations such that (1) the balance never drops below zero and (2) the
\cf{getBalance} operation never incorrectly returns a negative balance. To be
able to do this, we need to consider the anomalies that could arise under
eventual consistency.

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification over \cf{withdraw}:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}

\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

\noindent For \cf{getBalance}, we construct the following contract:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}

\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

@@

Since there are no restrictions on when and how a \cf{deposit} operation can
execute, its contract is simply $\small \true$.
